Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-21 talk/pepm2011/paper.tex
did those
some more tweaks by stephan

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
using partial evaluation. The use of partial evaluation is a bit peculiar in
that it receives no static input arguments for the trace, but it is only used to
optimize operations within the trace. This section will give an informal account
of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
line numbers are the lines of the unoptimized trace where the operation comes
from).

To optimize the trace, it is traversed from beginning to end and an output trace
is produced at the same time. Every operation in the input trace is either
removed, or put into the output trace. Sometimes new operations need to be
produced as well. The optimizer can only remove operations that manipulate
objects that have been allocated within the trace, while all other operations are copied to the
output trace untouched.

Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
in lines 1-9 are manipulating objects which existed before the trace and that
are passed in as arguments: therefore the optimizer just puts them into the
output trace.

The following operations (lines 10-17) are more interesting:
% XXX line numbers

@@

The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:

@@

The \lstinline{guard_class} operations can be removed, since their argument is a
static object with the matching type \lstinline{BoxedInteger}. The
\lstinline{get} operations can be removed as well, because each of them reads a
field out of a static object. The results of the get operation are replaced with
what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
replaced by $i_{4}$ and -100. The only operation put into the optimized trace
is the addition:

@@

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
similar vein. The operations in lines 27-35 produce two more static objects and
are removed. Those in line 36-39 are just put into the output trace because they
manipulate objects that are allocated before the trace. Lines 40-42 are removed
because they operate on a static object. Line 43 is put into the output trace.
Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
that static object and are removed as well. Lines 52-54 are put into the output
trace.

The last operation (line 55) is an interesting case. It is the \lstinline{jump}
operation that passes control back to the beginning of the trace. The two
arguments to this operation are at this point static objects. However, because
they passed into the next iteration of the loop they live longer than the trace
and therefore cannot remain static. They need to be turned into a dynamic
(runtime) object before the actual \lstinline{jump} operation. This process of
turning a static object into a dynamic one is called \emph{lifting}.

Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
the output trace. Those operations produce an object at runtime that has the
same shape that the static object describes. This process is a bit delicate,
because the static objects could form an arbitrary graph structure. In our
example it is simple, though:

@@

A bit more generally, lifting needs to occur if a static object is used in any
operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
It also needs to occur if \lstinline{set} is used to store a static object into
a non-static one.

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven.

@@

XXX input/output of optimizer

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. The environment of the
optimizer serves a function similar to that of the environment in the
semantics: to express sharing.

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. In addition to the run times the table also reports the
speedup that PyPy achieves when the optimization is turned on. XXX sounds sucky

>>>>

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
using partial evaluation. The use of partial evaluation is a bit peculiar in
that it receives no static input arguments for the trace, but it is only used to
optimize operations within the trace. This section will give an informal account
of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
line numbers are the lines of the unoptimized trace where the operation originates).

To optimize the trace, it is traversed from beginning to end while an output
trace is produced. Every operation in the input trace is either
removed or put into the output trace. Sometimes new operations need to be
produced as well. The optimizer can only remove operations that manipulate
objects that have been allocated within the trace, while all other operations are copied to the
output trace unchanged.

Looking at the example trace of Figure~\ref{fig:unopt-trace}, the operations
in lines 1--9 are manipulating objects which existed before the trace and that
are passed in as arguments: therefore the optimizer just puts them into the
output trace.

The following operations (lines 10--17) are more interesting:

@@

The subsequent operations (line 20--26) in Figure~\ref{fig:unopt-trace}, which
use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:

@@

The \lstinline{guard_class} operations can be removed, since their argument is a
static object with the matching type \lstinline{BoxedInteger}. The
\lstinline{get} operations can be removed as well, because each of them reads a
field out of a static object. The results of the get operation are replaced with
what the static object stores in these fields: all the occurences of $i_{7}$ and $i_{8}$ in the trace are just
replaced by $i_{4}$ and -100. The only operation copied into the optimized trace
is the addition:

@@

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
similar vein. The operations in lines 27--35 produce two more static objects and
are removed. Those in line 36--39 are just put into the output trace because they
manipulate objects that are allocated before the trace. Lines 40--42 are removed
because they operate on a static object. Line 43 is put into the output trace.
Lines 44--46 produce a new static object and are removed, lines 48--51 manipulate
that static object and are removed as well. Lines 52--54 are put into the output
trace.

The last operation (line 55) is an interesting case. It is the \lstinline{jump}
operation that passes control back to the beginning of the trace. The two
arguments to this operation at this point are static objects. However, because
they are passed into the next iteration of the loop they live longer than the trace
and therefore cannot remain static. They need to be turned into dynamic
(runtime) objects before the actual \lstinline{jump} operation. This process of
turning a static object into a dynamic one is called \emph{lifting}.

Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
the output trace. Those operations produce an object at runtime that has the
shape described by the static object. This process is a bit delicate,
because the static objects could form an arbitrary graph structure. In our
example it is simple, though:

@@

More generally, lifting needs to occur if a static object is used in any
operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
It also needs to occur if \lstinline{set} is used to store a static object into
a non-static one.

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original
five, and only three \lstinline{guard_class} operations, compared to the
original seven.

@@

The state of the optimizer is stored in an environment $E$ and a \emph{static
heap} $S$. Each step of the optimizer takes an operation, an environment and a
static heap and produces a list of operations, a new environment and a new
static heap.

The environment is a partial function from variables in the
unoptimized trace $V$ to variables in the optimized trace $V^*$ (which are
themselves written with a
$\ ^*$ for clarity). The reason for introducing new variables in the optimized
trace is that several variables that appear in the unoptimized trace can turn
into the same variables in the optimized trace. The environment of the
optimizer serves a function similar to that of the environment in the
semantics: to express sharing.

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. For each implementation the table also reports the
speedup that PyPy with optimization achieves over it.
